# Dataframes {#sec-dataframes}

```{r}
#| echo: false
library(tidyverse)
library(fivethirtyeight)
```

## What is a tibble?

## Converting to "tidy" data

For the rest of this book, we will primarily deal with data that is already in "tidy" format as well. This is not always the case however with all datasets in the world. If your original data frame is in wide (non-"tidy") format and you would like to use the tidyverse packages, you will first have to convert it to "tidy" format. To do so, we recommend using the \index{tidyr!pivot\_longer()} `pivot_longer()` function in the `tidyr` \index{R packages!tidyr} package [@R-tidyr].


Let's load some data from the `fivethirtyeight` package. The `fivethirtyeight` package [@R-fivethirtyeight] provides access to the datasets used in many articles published by the data journalism website, [FiveThirtyEight.com](https://fivethirtyeight.com/). For a complete list of all `r nrow(data(package = "fivethirtyeight")[[3]])` datasets included in the `fivethirtyeight` package, check out the package webpage by going to: <https://fivethirtyeight-r.netlify.app/articles/fivethirtyeight.html>.\index{R packages!fivethirtyeight}

Let's focus our attention on the `drinks` data frame:

```{r}
#| echo: false
drinks %>%
  head(5)
```

After reading the help file by running (`?drinks`), you'll see that `drinks` is a data frame containing results from a survey of beer, spirits, and wine consumption originally reported on FiveThirtyEight.com in Mona Chalabi's article: ["Dear Mona Followup: Where Do People Drink The Most Beer, Wine And Spirits?"](https://fivethirtyeight.com/features/dear-mona-followup-where-do-people-drink-the-most-beer-wine-and-spirits/).

Let's narrow down the data a bit.  We'll only consider 4 countries (the United States, China, Italy, and Saudi Arabia), omit the `total_litres_of_pure_alcohol` variable, and rename the other variables to something a bit more convenient.  Don't worry about the code here.  We'll get into all of these operations more in the chapter on @sec-dplyr.

```{r}
drinks_smaller <- drinks %>% 
  filter(country %in% c("USA", "China", "Italy", "Saudi Arabia")) %>% 
  select(-total_litres_of_pure_alcohol) %>% 
  rename(beer = beer_servings, spirit = spirit_servings, wine = wine_servings)
drinks_smaller
```

Note that this data is not in "tidy" format.  However, we can convert it to tidy format by using the `pivot_longer()` function from the `tidyr` package as follows:

```{r}
drinks_smaller_tidy <- drinks_smaller %>% 
  pivot_longer(names_to = "type", 
               values_to = "servings", 
               cols = -country)
drinks_smaller_tidy
```

```{r}
#| echo: false
n_row_drinks <- drinks_smaller_tidy %>% nrow()
n_alcohol_types <- drinks_smaller_tidy %>%
  select(type) %>%
  n_distinct()
n_countries <- drinks_smaller_tidy %>%
  select(country) %>%
  n_distinct()
```

We set the arguments to `pivot_longer()` as follows:

1. `names_to` here corresponds to the name of the variable in the new "tidy"/long data frame that will contain the *column names* of the original data. Observe how we set `names_to = "type"`. In the resulting `drinks_smaller_tidy`, the column `type` contains the three types of alcohol `beer`, `spirit`, and `wine`. Since `type` is a variable name that doesn't appear in `drinks_smaller`, we use quotation marks around it. You'll receive an error if you just use `names_to = type` here.
1. `values_to` here is the name of the variable in the new "tidy" data frame that will contain the *values* of the original data. Observe how we set `values_to = "servings"` since each of the numeric values in each of the `beer`, `wine`, and `spirit` columns of the `drinks_smaller` data corresponds to a value of `servings`. In the resulting `drinks_smaller_tidy`, the column `servings` contains the `r n_countries` $\times$ `r n_alcohol_types` = `r n_row_drinks` numerical values. Note again that `servings` doesn't appear as a variable in `drinks_smaller` so it again needs quotation marks around it for the `values_to` argument.
1. The third argument `cols` is the columns in the `drinks_smaller` data frame you either want to or don't want to "tidy." Observe how we set this to `-country` indicating that we don't want to "tidy" the `country` variable in `drinks_smaller` and rather only `beer`, `spirit`, and `wine`. Since `country` is a column that appears in `drinks_smaller` we don't put quotation marks around it.

The third argument here of `cols` is a little nuanced, so let's consider code that's written slightly differently but that produces the same output: 

```{r}
#|eval: false
drinks_smaller %>% 
  pivot_longer(names_to = "type", 
               values_to = "servings", 
               cols = c(beer, spirit, wine))
```

Note that the third argument now specifies which columns we want to "tidy" with `c(beer, spirit, wine)`, instead of the columns we don't want to "tidy" using `-country`. We use the `c()` function to create a vector of the columns in `drinks_smaller` that we'd like to "tidy." Note that since these three columns appear one after another in the `drinks_smaller` data frame, we could also do the following for the `cols` argument:

```{r}
drinks_smaller %>% 
  pivot_longer(names_to = "type", 
               values_to = "servings", 
               cols = beer:wine)
drinks_smaller_tidy <- drinks_smaller %>%
  gather(type, servings, -country)
```

Let's see what our "tidy" formatted data frame, `drinks_smaller_tidy`, looks like.

```{r}
#| echo: false
drinks_smaller_tidy
```

Converting "wide" format data to "tidy" format often confuses new R users. The only way to learn to get comfortable with the `pivot_longer()` function is with practice, practice, and more practice using different datasets. For example, run `?pivot_longer` and look at the examples in the bottom of the help file.

If however you want to convert a "tidy" data frame to "wide" format, you will need to use the `pivot_wider()`\index{tidyr!pivot\_wider()} function instead. Run `?pivot_wider` and look at the examples in the bottom of the help file for examples.

You can also view examples of both `pivot_longer()` and `pivot_wider()` on the [tidyverse.org](https://tidyr.tidyverse.org/dev/articles/pivot.html#pew) webpage. There's a nice example to check out the different functions available for data tidying and a case study using data from the World Health Organization on that webpage. Furthermore, each week the R4DS Online Learning Community posts a dataset in the weekly [`#`TidyTuesday event](https://github.com/rfordatascience/tidytuesday) that might serve as a nice place for you to find other data to explore and transform. 